Finiteness in Cubical Type Theory

### 2020-06-15 sections/introduction.tex
more converted into agda notation

Finally, we will also want to optimise our solutions and solver: for this we
will remove equivalent expressions, which can be accomplished with quotients.
\section{Notation and Background}
\subsection{Notation}
We work in Cubical Type Theory \cite{cohenCubicalTypeTheory2016}.
For the various type formers we use the following notation:
\subsubsection{Type}
We use \(\AgdaDatatype{Type}\) to denote the universe of (small) types.
The universe level is denoted with a subscript number, starting at 0.
``Type families'' are functions into \(\AgdaDatatype{Type}\).
\subsubsection{Simple Types}
The three basic types---often called 0, 1, and 2---in MLTT here will be denoted
with their more common names: \(\bot\), \(\top\), and \(\mathbf{Bool}\),
respectively.
\begin{figure}[h!]
  \begin{multicols}{3}
    \begin{agdalisting}
      \ExecuteMetaData[agda/Snippets/Introduction.tex]{bot}
    \end{agdalisting}
    \begin{agdalisting}
      \ExecuteMetaData[agda/Snippets/Introduction.tex]{top}
    \end{agdalisting}
    \begin{agdalisting}
      \ExecuteMetaData[agda/Snippets/Introduction.tex]{bool}
    \end{agdalisting}
  \end{multicols}
  \caption{Agda Definitions of the Three Basic Types}
\end{figure}
\subsubsection{Dependent Sums}
Dependent sums are denoted with the usual \(\Sigma\) symbol, and are defined in
Fig.~\ref{sigma-def}.
\begin{figure}[h!]
  \begin{multicols}{2}
    \begin{agdalisting}
      \ExecuteMetaData[agda/Snippets/Introduction.tex]{sigma}
    \end{agdalisting}
    \begin{agdalisting}
      \ExecuteMetaData[agda/Snippets/Introduction.tex]{sigma-syntax-1}
      \ExecuteMetaData[agda/Snippets/Introduction.tex]{sigma-syntax-2}
      \ExecuteMetaData[agda/Snippets/Introduction.tex]{sigma-syntax-3}
    \end{agdalisting}
  \end{multicols}
  \caption{Agda Definitions of Dependent Sums}
  \label{sigma-def}
\end{figure}

\begin{description}
  \item[0 , 1 , 2] We call the \(\mathbf{0}\), \(\mathbf{1}\), and
    \(\mathbf{2}\) types \(\bot\), \(\top\), and \(\mathbf{Bool}\) respectively.
    The single inhabitant of \(\top\) is tt, and the two inhabitants of
    \(\mathbf{Bool}\) are false and true.
    The ``negation'' of a type, written \(\neg A\), means \(A \rightarrow
    \bot\).
  \item[Dependent Sum and Product] We use \(\Sigma\) and \(\Pi\) for the
    dependent sum and product, respectively.
    The two projections from \(\Sigma\) are called fst and snd.
    In the non-dependent case, \(\Sigma\) can be written as \(\times\), and
    \(\Pi\) as \(\rightarrow\).
  \item[Disjoint Union] We define disjoint union as an inductive type.
    \begin{equation}
      \begin{alignedat}{3}
        A \uplus B \coloneqq & \;
        \text{inl} &: A \rightarrow A \uplus B ; \\
        | & \;  \text{inr} &: B \rightarrow A \uplus B ;
      \end{alignedat}
    \end{equation}
    It is also expressible with only \(\Sigma\):
    \(A \uplus B \simeq \Sigma(x : \mathbf{Bool}) , \text{if } x \text{ then
    } A \text{ else } B \).
  \item[Equalities, equivalences, and paths] We use the symbol \(\coloneqq\)
    for definitions.
    \(\simeq\) will be used for equivalences, and \(\equiv\) for equalities.
    Of course, we know that \((A \simeq B) \simeq (A \equiv B)\) by univalence,
    so the distinction isn't terribly important in usage: we will only use one
    or the other as a suggestion of how we constructed it or how it is to be
    used.
  \item[Lambdas] \todo{Figure out how to describe lambdas}
\end{description}
\subsection{Cubical Type Theory}
Cubical Type Theory \cite{cohenCubicalTypeTheory2016} is a constructive type
theory with an implementation in Cubical Agda
\cite{vezzosiCubicalAgdaDependently2019}.
It allows us to do much of the same theory as in HoTT, but crucially the
univalence ``axiom'' is a \emph{theorem}, giving it computational content.
\begin{definition}[Path Types] \label{path-types}
  The equality type (which we denote with \(\equiv\)) in CuTT is the type of
  Paths\footnotemark.
  The internal structure of paths is largely irrelevant to us here, as we will
  generally treat \(\equiv\) as a black-box equivalence relation with
  substitution and congruence.
\end{definition}

>>>>

Finally, we will also want to optimise our solutions and solver: for this we
will remove equivalent expressions, which can be accomplished with quotients.
\section{Notation and Background}
We work in Cubical Type Theory \cite{cohenCubicalTypeTheory2016}.
The notation we use comes from the conventions used in the Agda standard
library.
\begin{definition}[Type]
  We use \(\AgdaDatatype{Type}\) to denote the universe of (small) types.
  The universe level is denoted with a subscript number, starting at 0.
  ``Type families'' are functions into \(\AgdaDatatype{Type}\).
\end{definition}
\begin{definition}[Basic Types]
  The three basic types---often called 0, 1, and 2 in MLTT---here will be
  denoted with their more common names: \(\bot\), \(\top\), and
  \(\mathbf{Bool}\), respectively.
  \begin{multicols}{3} \centering
    \begin{agdalisting}
      \ExecuteMetaData[agda/Snippets/Introduction.tex]{bot}
    \end{agdalisting}

    \begin{agdalisting}
      \ExecuteMetaData[agda/Snippets/Introduction.tex]{top}
    \end{agdalisting}

    \begin{agdalisting}
      \ExecuteMetaData[agda/Snippets/Introduction.tex]{bool}
    \end{agdalisting}
  \end{multicols}
\end{definition}
\begin{definition}[The Dependent Sum]
  Dependent sums are denoted with the usual \(\Sigma\) symbol, and has the
  following definition in Agda:

  \begin{center}
    \begin{agdalisting}
      \ExecuteMetaData[agda/Snippets/Introduction.tex]{sigma}
    \end{agdalisting}
  \end{center}
  We will use different notations to refer to this type depending on the
  setting.
  The following four expressions all denote the same type:

  \begin{multicols}{4} \centering
    \begin{agdalisting}
      \ExecuteMetaData[agda/Snippets/Introduction.tex]{sigma-syntax-1}
    \end{agdalisting}

    \begin{agdalisting}
      \ExecuteMetaData[agda/Snippets/Introduction.tex]{sigma-syntax-2}
    \end{agdalisting}

    \begin{agdalisting}
      \ExecuteMetaData[agda/Snippets/Introduction.tex]{sigma-syntax-3}
    \end{agdalisting}

    \begin{agdalisting}
      \ExecuteMetaData[agda/Snippets/Introduction.tex]{sigma-syntax-4}
    \end{agdalisting}
  \end{multicols}

  The non-dependent product is a special instance of the dependent.
  We indicate a simple pair of types \(A\) and \(B\) as \(A \times B\).
\end{definition}
\begin{definition}[Dependent Product]
  Dependent products (dependent functions) use the \(\Pi\) symbol.

  The three following expressions all denote the same type:

  \begin{multicols}{3} \centering
    \begin{agdalisting}
      \ExecuteMetaData[agda/Snippets/Introduction.tex]{pi-syntax-1}
    \end{agdalisting}

    \begin{agdalisting}
      \ExecuteMetaData[agda/Snippets/Introduction.tex]{pi-syntax-2}
    \end{agdalisting}

    \begin{agdalisting}
      \ExecuteMetaData[agda/Snippets/Introduction.tex]{pi-syntax-3}
    \end{agdalisting}
  \end{multicols}

  Again, the non-dependent case is a special case of the dependent case.
  Non-dependent functions are denoted with the arrow (\(\rightarrow\)).
\end{definition}

\begin{description}
  \item[Disjoint Union] We define disjoint union as an inductive type.
    \begin{equation}
      \begin{alignedat}{3}
        A \uplus B \coloneqq & \;
        \text{inl} &: A \rightarrow A \uplus B ; \\
        | & \;  \text{inr} &: B \rightarrow A \uplus B ;
      \end{alignedat}
    \end{equation}
    It is also expressible with only \(\Sigma\):
    \(A \uplus B \simeq \Sigma(x : \mathbf{Bool}) , \text{if } x \text{ then
    } A \text{ else } B \).
  \item[Equalities, equivalences, and paths] We use the symbol \(\coloneqq\)
    for definitions.
    \(\simeq\) will be used for equivalences, and \(\equiv\) for equalities.
    Of course, we know that \((A \simeq B) \simeq (A \equiv B)\) by univalence,
    so the distinction isn't terribly important in usage: we will only use one
    or the other as a suggestion of how we constructed it or how it is to be
    used.
  \item[Lambdas] \todo{Figure out how to describe lambdas}
\end{description}
\subsection{Cubical Type Theory}
Cubical Type Theory \cite{cohenCubicalTypeTheory2016} is a constructive type
theory with an implementation in Cubical Agda
\cite{vezzosiCubicalAgdaDependently2019}.
It allows us to do much of the same theory as in HoTT, but crucially the
univalence ``axiom'' is a \emph{theorem}, giving it computational content.
\begin{definition}[Path Types] \label{path-types}
  The equality type (which we denote with \(\equiv\)) in CuTT is the type of
  Paths\footnotemark.
  The internal structure of paths is largely irrelevant to us here, as we will
  generally treat \(\equiv\) as a black-box equivalence relation with
  substitution and congruence.
\end{definition}
